A Sound Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2022-06-21 paper/paper.tex
Small textual fixes

Although there are some dedicated provers for modal logics~\cite{Gleissner17,Mora11},
we believe that using mature tools, like a proof assistant, are a better option for
verifying facts using modal logics. We justify this choice by: 1) the recent
successful histories on verification of software and
mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) There is a growing
community of developers using such tools thanks to 
textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
newbies to join the game of formal verification and 3) The custom automation
facilities supported by current proof assistants allows that trivial facts can
be proved using decision procedures, relieving the user from writing proof
scripts for formulas that can be checked automatically.

@@

In this section, we describe the implementation details of our modal logic
library in Coq. We start by describing the structure of formula syntax and the
representation of frames and models, needed to define the Kripke semantics.
Next, we define a modular Hilbert-style deductive system. The modularization is
obtained by parameterizing it by a set of axioms used. This allows us to
generalize results about derivations to several deductive systems, independently
of the axioms used. We conclude by discussing the soundness, the current
status of the completeness and the frame correspondence proofs.

@@

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of on four
possibilities in a proof: 1) the use of a hypothesis; 2) use of an axiom;
3) use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
axioms that form a system for modal logics. In this way, our definition can be
used by several systems of modal logic.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop)
                           : theory -> form -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: form)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: form),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: form)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: form)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discuss the rule for axioms, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom : Set :=
  | ax1   : form -> form -> axiom
  | ax2   : form -> form -> form  -> axiom
  (** some code omitted *)
  | axK   : form -> form -> axiom
  | axPos : form -> form -> axiom
  | axT   : form -> axiom
  | axB   : form -> axiom
  | axK4  : form -> axiom
  | axD   : form -> axiom
  | axK5  : form -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 consists encode axioms for propositional logics
and the last 7 are for different formulas taken as axioms by systems of modal logics.
Axioms are built using the function \mintinline{coq}{instantiate}, which
receives an axiom and build its corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): modalFormula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (** some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

The \mintinline{coq}{Ax} constructor uses the \mintinline{coq}{instantiate}
function to create a formula instance from using a given system,
represented as a predicate
\mintinline{coq}{A : axiom -> Prop}, which parameterizes our deduction type. An
axiom \mintinline{coq}{a} is valid within a system \mintinline{coq}{A} whenever
the type \mintinline{coq}{A a} is inhabited, i.e. it is not empty. Next, we
detail the representation of different systems of modal logic in our library.

\paragraph{Systems of modal logics} In order to use the deductive system to
check the provability of formulas, we need to specify the axioms that are
included in the system. Our approach is to define predicates
which denote each system. The simplest set of axioms is usually refereed as
\textbf{K} and it includes the axiom $\Box(A \to B) \to (\Box A \to \Box B)$
and all the axioms for propositional logic. We represent system \textbf{K}
by the following inductive type:
\begin{minted}{coq}
Inductive K: axiom -> Prop :=
  | K_ax1: forall p q, K (ax1 p q)
  | K_ax2: forall p q r, K (ax2 p q r)
  (** some code omitted *)
  | K_axK: forall p q, K (axK p q)
\end{minted}
Type \mintinline{coq}{K} is a predicate over axioms: it has one constructor for
each of the valid axioms which form the system. Given a value
of \mintinline{coq}{axiom}, a value of type \mintinline{coq}{K a} denotes a
proof that a given axiom is part of the system \textbf{K}. This is the main
trick used by the constructor \mintinline{coq}{Ax}, of type
\mintinline{coq}{deduction}, to generically represent axioms for
different systems of modal logics uniformly. In order to understand how it
works, let us review its type (below, remeber that \mintinline{coq}{A : axiom -> Prop}
denotes a system of axioms for modal logics):
\begin{minted}{coq}
Ax: forall (t: theory)
           (a: axiom)
           (f: form),
           A a -> instantiate a = f -> deduction A t f
\end{minted}
Notice that the first parameter of constructor \mintinline{coq}{Ax} is a value
of type \mintinline{coq}{A a}, which is only inhabited if the predicate
\mintinline{coq}{A} holds for the axiom \mintinline{coq}{a}. Through this
evidence, is possible to create the corresponding axiom formula instance by
using function \mintinline{coq}{instantiate}. In this way, we are able to
represent all systems of axioms for modal logics using a single constructor
for its axioms.

>>>>

Although there are some dedicated provers for modal logics~\cite{Gleissner17,Mora11},
we believe that using mature tools, like a proof assistant, is a better option for
verifying facts using modal logics. We justify this choice by: 1) the recent
successful histories on verification of software and
mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) There is a growing
community of developers using such tools thanks to 
textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
newbies to join the game of formal verification and 3) The custom automation
facilities supported by current proof assistants allows that trivial facts can
be proved using decision procedures, relieving the user from writing proof
scripts for formulas that can be checked automatically.

@@

In this section, we describe the implementation details of our modal logic
library in Coq. We start by describing the structure of formula syntax and the
representation of frames and models, needed to define the Kripke semantics.
Next, we define a modular Hilbert-style deductive system. The modularization is
obtained by parameterizing it by a set of axioms used. This allows us to
generalize results about derivations to several deductive systems, independently
of the axioms used. We conclude by discussing the soundness and the frame
correspondence proofs.

@@

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows using one out of four
possibilities in a proof: 1) the use of a hypothesis; 2) use of an axiom;
3) use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
axioms that form a system for modal logics. In this way, our definition can be
used by several systems of modal logic.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop)
                           : theory -> form -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: form)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: form),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: form)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: form)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discuss the \mintinline{coq}{Ax}, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom : Set :=
  | ax1   : form -> form -> axiom
  | ax2   : form -> form -> form  -> axiom
  (** some code omitted *)
  | axK   : form -> form -> axiom
  | axPos : form -> form -> axiom
  | axT   : form -> axiom
  | axB   : form -> axiom
  | axK4  : form -> axiom
  | axD   : form -> axiom
  | axK5  : form -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 consists encode axioms for propositional logics
and the last 7 are for different formulas taken as axioms by systems of modal logics.
Axioms are built using the function \mintinline{coq}{instantiate}, which
receives an axiom and build its corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): modalFormula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (** some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

The \mintinline{coq}{Ax} constructor uses the \mintinline{coq}{instantiate}
function to create a formula instance from a given system,
represented as a predicate
\mintinline{coq}{A : axiom -> Prop} which parameterizes our deduction type. An
axiom \mintinline{coq}{a} is valid within a system \mintinline{coq}{A} whenever
the type \mintinline{coq}{A a} is inhabited, i.e. it is not empty. Next, we
describe how to represent different systems of modal logic.

\paragraph{Systems of modal logics} In order to use the deductive system to
check the provability of formulas, we need to specify the axioms that are
included in the system. Our approach is to define predicates
which denote each system. The simplest set of axioms is usually refereed as
\textbf{K} and it includes the axiom $\Box(A \to B) \to (\Box A \to \Box B)$
and all the axioms for propositional logic. We represent system \textbf{K}
by the following inductive type:
\begin{minted}{coq}
Inductive K: axiom -> Prop :=
  | K_ax1: forall p q, K (ax1 p q)
  | K_ax2: forall p q r, K (ax2 p q r)
  (** some code omitted *)
  | K_axK: forall p q, K (axK p q)
\end{minted}
Type \mintinline{coq}{K} is a predicate over axioms: it has one constructor for
each of the valid axioms which form the system. Given a value
of \mintinline{coq}{axiom}, a value of type \mintinline{coq}{K a} denotes a
proof that a given axiom is part of the system \textbf{K}. This is the main
trick used by the constructor \mintinline{coq}{Ax}, of type
\mintinline{coq}{deduction}, to generically represent axioms for
different systems of modal logics uniformly. In order to understand how it
works, let us review its type (below, remember that \mintinline{coq}{A : axiom -> Prop}
denotes a system of axioms for modal logics):
\begin{minted}{coq}
Ax: forall (t: theory)
           (a: axiom)
           (f: form),
           A a -> instantiate a = f -> deduction A t f
\end{minted}
Note that the first parameter of constructor \mintinline{coq}{Ax} is a value
of type \mintinline{coq}{A a}, which is only inhabited if the predicate
\mintinline{coq}{A} holds for the axiom \mintinline{coq}{a}. Through this
evidence, is possible to create the corresponding axiom formula instance by
using function \mintinline{coq}{instantiate}. In this way, we are able to
represent all systems of axioms for modal logics using a single constructor
for its axioms.
